MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-13 docs/paper/out/paper.tex
put Materials & Methods *after* Discussion -- paper
seriously restructure the paper structure -- paper
Much better! it starts to look like a paper.

During the last years other sciences as astronomy or particle physics
are facing the big data challenge but, at least, these science have
standards for data processing \citep{stephens2015big}. Global standards
for converting raw sequence data into processed data are not yet well
defined in metagenomics and there are shortcomings derived from the fact
that many bioinformatics methodologies currently used for metagenomics
data analysis were designed for a scenario very different that the
current one. These are some of the aspects that have suffered crucial
changes and advances with a direct impact in metagenomics data analysis.
i. The first aspect is related to the sequences to be analyzed: the
reads are larger, the sequencing depth and the number of samples of each
project are considerably bigger. The first metagenomics studies were
very local projects, while nowadays the most fruitful studies are done
at a global level (international, continental, national). This kind of
global studies has yielded the discovery of clinical biomarkers for
diseases of the importance of cancer, obesity or inflammatory bowel
diseases and has allowed exploring the biodiversity in many earth
environments ii. The second aspect derives from the impressive genomics
explosion, its effect being felt in this case in the reference
sequences. The immense amount of sequences available in public
repositories demands new approaches in curation, update and storage for
metagenomics reference databases: current models will or already have
problems to face the future avalanche of metagenomic sequences. iii. The
third aspect to consider for metagenomics data analysis is related to
the appearance of new models for massive computation and storage and to
the new programming methodologies (Scala, \ldots{}) and new cloud models
and resources. The immense new possibilities that these advances offer
must have a direct impact in the metagenomics data analysis. iv. And
finally the new social manner to do science, and especially genomic
science is the fourth aspect to consider. Metagenomics evolves in a
social and global scenario following a science democratization trend in
which many small research groups from distant countries share a common
big metagenomics project. This global cooperation demands systems
allowing following exactly the same pipelines using equivalent cloud
resources to modularly execute the analysis in an asynchronous way of
working between different groups. This definitively new scenario demands
new methods and tools to handle the current and future volume of
metagenomic data with the sufficient speed of analysis. Considering all
these aspects we have designed a new open source methodology for
analyzing metagenomics data that exploits the new possibilities that
cloud computing offers to get a system robust, programmatically
configurable, modular, distributed, flexible, scalable and traceable in
which the biological databases of reference sequences can be easily
updated and/or frequently substituted by new ones or by databases
specifically designed for focused projects.

\href{http://www.scala-lang.org/}{Scala} is a hybrid object-functional
programming language which runs on Java Virtual Machine. It has support
for type-level programming, type-dependent types (through type members)
and singleton types, which permits a restricted form of dependent types
where types can depend essentially on values determined at compile time
(through their corresponding singleton types). Conversely, through
implicits one can retrieve the value corresponding to a singleton type.

The other key feature for us is Java interoperability, which let us
build on the vast number of existing Java libraries; we take advantage
of this when using Bio4j as an API for the NCBI taxonomy.

@@

MG7 uses
\href{https://github.com/statika/statika/releases/tag/v0.2.0}{ohnosequences/statika
2.0.0}.

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors to declare datasets and
their locations. \textbf{Data} is represented as type-indexed fields:
Keys are modeled as singleton types, and values correspond to what could
be called a denotation of the key: a value of type \texttt{Location}
tagged with the key type. Then a \textbf{Dataset} is essentially a
collection of data, which are guaranteed statically to be different
through type-level predicates, making use of the value ↔ type
correspondence which can be established through singleton types and
implicits. A dataset location is then just a list of locations formed by
locations of each data member of that dataset. All this is based on what
could be described as an embedding in Scala of an extensible record
system with concatenation on disjoint labels, in the spirit of
\citep[\citet{harper1991record}]{harper1990extensible}. For that
\emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/releases/tag/v0.7.0}{ohnosequences/cosas
0.7.0}.

@@

MG7 uses
\href{https://github.com/ohnosequences/datasets/releases/tag/v0.2.0}{ohnosequences/datasets
0.2.0}.

@@

MG7 uses
\href{https://github.com/ohnosequences/loquat/releases/tag/v2.0.0}{ohnosequences/loquat
2.0.0}.

\subsection{Type-safe EDSLs for BLAST and
FLASH}\label{type-safe-edsls-for-blast-and-flash}

We developed our own Scala-based type-safe EDSLs (Embedded Domain
Specific Language) for
\href{https://github.com/ohnosequences/flash}{FLASH} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

\subsubsection{BLAST EDSL}\label{blast-edsl}

@@

MG7 uses
\href{https://github.com/ohnosequences/blast/releases/tag/v0.2.0}{ohnosequences/blast
0.2.0}.

\subsubsection{FLASH EDSL}\label{flash-edsl}

@@

MG7 uses
\href{https://github.com/ohnosequences/flash/releases/tag/v0.1.0}{ohnosequences/flash
0.1.0}.

\href{https://github.com/bio4j/bio4j}{Bio4j} is a data platform
integrating data from different resources such as UniProt or GO in a
graph data paradigm. We use the module containing the NCBI Taxonomy, and
the use their Java API from Scala in the assignment phase.

MG7 uses
\href{https://github.com/bio4j/bio4j/releases/tag/v0.12.0-RC3}{bio4j/bio4j
0.12.0-RC3} and
\href{https://github.com/bio4j/bio4j-titan/releases/tag/v0.4.0-RC2}{bio4j/bio4j-titan
0.4.0-RC2}.

To tackle the challenges posed by metagenomics big data analysis
outlined in the Introduction,

­- AWS resources in Scala (??) - A new approach to data analysis
specification, management and specification based on working with it in
exactly the same way as for a software project, together with the
extensive use of compile-time structures and checks. - Parallelization
and distributed analysis based on AWS, with on-demand infrastructure as
the basic paradigm ­- fully automated processes, data and cloud
resources management. - Static reproducible specification of
dependencies and behavior of the different components using
\emph{Statika} and \emph{Datasets} - Definition of complex pipelines
using \emph{Loquat} a composable system for scaling/parallelizing
stateless computations especially designed for Amazon Web Services (AWS)
- Modeling of the taxonomy tree using the new paradigm of graph
databases (Bio4j). It facilitates the taxonomic assignment tasks and the
calculation of the taxa abundance values considering the hierarchic
structure of taxonomy tree (cumulative values). - per-read assignment
(??)

\subsection{16S Reference Database
Construction}\label{s-reference-database-construction}

@@

\subsection{Bio4j and Graph Databases}\label{bio4j-and-graph-databases}

\subsection{MG7 Pipeline Description}\label{mg7-pipeline-description}

\subsection{Taxonomic Assignment
Algorithms}\label{taxonomic-assignment-algorithms}

\subsubsection{Lowest Common Ancestor based Taxonomic
Assignment}\label{lowest-common-ancestor-based-taxonomic-assignment}

@@

\subsubsection{Best BLAST hit taxonomic
assignment}\label{best-blast-hit-taxonomic-assignment}

@@

\subsection{MG7 availability}\label{mg7-availability}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Treat data analysis as a software project. This makes for radical
  improvements in \emph{reproducibility}, \emph{reuse},
  \emph{versioning}, \emph{safety}, \emph{automation} and
  \emph{expressiveness}
\item
  input and output data, their locations and type are expressible and
  checked at compile-time using our Scala library \emph{datasets} ­3.
  management of dependencies and machine configurations using our Scala
  library \emph{Statika}
\item
  automation of AWS cloud resources and processes, including
  distribution and parallelization through the use of \emph{Loquat}
\item
  taxonomic data and related operations are treated natively as what
  they are: graphs, through the use of \emph{Bio4j}
\item
  MG7 provides a sustainable model for taxonomic assignment, appropriate
  to face the challenging amount of data that high throughput sequencing
  technologies generate
\end{enumerate}

@@

Other possible uses of the general schema: statika, loquat, \ldots{}

@@

@@

INTERCROSSING (Grant 289974)

>>>>

During the last years other sciences as astronomy or particle physics
are facing the big data challenge but, at least, these science have
standards for data processing \citep{stephens2015big}. Global standards
for converting raw sequence data into processed data are not yet well
defined in metagenomics and there are shortcomings derived from the fact
that many bioinformatics methodologies currently used for metagenomics
data analysis were designed for a scenario very different that the
current one. These are some of the aspects that have suffered crucial
changes and advances with a direct impact in metagenomics data analysis:

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  The first aspect is related to the sequences to be analyzed: the reads
  are larger, the sequencing depth and the number of samples of each
  project are considerably bigger. The first metagenomics studies were
  very local projects, while nowadays the most fruitful studies are done
  at a global level (international, continental, national). This kind of
  global studies has yielded the discovery of clinical biomarkers for
  diseases of the importance of cancer, obesity or inflammatory bowel
  diseases and has allowed exploring the biodiversity in many earth
  environments
\item
  The second aspect derives from the impressive genomics explosion, its
  effect being felt in this case in the reference sequences. The immense
  amount of sequences available in public repositories demands new
  approaches in curation, update and storage for metagenomics reference
  databases: current models will or already have problems to face the
  future avalanche of metagenomic sequences.
\item
  The third aspect to consider for metagenomics data analysis is related
  to the appearance of new models for massive computation and storage
  such as the so-called cloud, or the widespread adoption of programming
  methodologies like functional programming, or, more speculatively,
  dependently typed programming. The immense new possibilities that
  these advances offer must have a direct impact in metagenomics data
  analysis.
\item
  And finally the new social manner to do science, and especially
  genomic science is the fourth aspect to consider. Metagenomics evolves
  in a social and global scenario following a science democratization
  trend in which many small research groups from distant countries share
  a common big metagenomics project. This global cooperation demands
  systems allowing following exactly the same pipelines using equivalent
  cloud resources to modularly execute the analysis in an asynchronous
  way of working between different groups. This new scenario calls for
  new methods and tools to handle the current and future volume of
  metagenomic data with the sufficient speed of analysis.
\end{enumerate}

Considering all these aspects we have designed a new open source
methodology for analyzing metagenomics data that exploits the new
possibilities that cloud computing offers to get a system robust,
programmatically configurable, modular, distributed, flexible, scalable
and traceable in which the biological databases of reference sequences
can be easily updated and/or frequently substituted by new ones or by
databases specifically designed for focused projects.

To tackle the challenges posed by metagenomics big data analysis
outlined in the Introduction

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using \emph{Statika} and \emph{Datasets}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex pipelines using \emph{Loquat}, a composable
  system for scaling/parallelizing stateless computations especially
  designed for Amazon Web Services (AWS)
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j). It facilitates the taxonomic assignment tasks and
  the calculation of the taxa abundance values considering the
  hierarchic structure of taxonomy tree (cumulative values)
\item
  per-read assignment (??)
\end{itemize}

\subsection{Libraries and resources}\label{libraries-and-resources}

In this section we describe the resources and Scala libraries developed
by the authors on top of which MG7 is built.

\href{http://www.scala-lang.org/}{Scala}, a hybrid object-functional
programming language, was chosen based on the possibility of using
certain advanced programming styles, and Java interoperability, which
let us build on the vast number of existing Java libraries; we take
advantage of this when using Bio4j as an API for the NCBI taxonomy. It
has support for type-level programming, type-dependent types (through
type members) and singleton types, which permits a restricted form of
dependent types where types can depend essentially on values determined
at compile time (through their corresponding singleton types).
Conversely, through implicits one can retrieve the value corresponding
to a singleton type.

\subsubsection{\texorpdfstring{\emph{Statika}: machine configuration and
behavior}{Statika: machine configuration and behavior}}\label{statika-machine-configuration-and-behavior}

@@

\subsubsection{\texorpdfstring{\emph{Datasets}: a mini-language for
data}{Datasets: a mini-language for data}}\label{datasets-a-mini-language-for-data}

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: Keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value ↔ type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each data member of
that dataset. All this is based on what could be described as an
embedding in Scala of an extensible record system with concatenation on
disjoint labels, in the spirit of
\citep[\citet{harper1991record}]{harper1990extensible}. For that
\emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}.

@@

\subsubsection{\texorpdfstring{\emph{Loquat}: Parallel data processing
with
AWS}{Loquat: Parallel data processing with AWS}}\label{loquat-parallel-data-processing-with-aws}

@@

\subsubsection{Type-safe eDSLs for BLAST and
FLASH}\label{type-safe-edsls-for-blast-and-flash}

We developed our own Scala-based type-safe eDSLs (\textbf{e}mbedded
\textbf{D}omain \textbf{S}pecific \textbf{L}anguage) for
\href{https://github.com/ohnosequences/flash}{FLASH} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

@@

@@

\subsubsection{Bio4j and Graph
Databases}\label{bio4j-and-graph-databases}

\citep[Bio4j][]{pareja2015bio4j} is a data platform integrating data
from different resources such as UniProt or GO in a graph data paradigm.
In the assignment phase we use a subgraph containing the NCBI Taxonomy,
wrapping in Scala its Java API in a tree algebraic data type.

\subsubsection{16S Reference Database
Construction}\label{s-reference-database-construction}

@@

\subsection{Pipeline Description}\label{pipeline-description}

\subsubsection{Taxonomic Assignment
Algorithms}\label{taxonomic-assignment-algorithms}

\paragraph{Lowest Common Ancestor based Taxonomic
Assignment}\label{lowest-common-ancestor-based-taxonomic-assignment}

@@

\paragraph{Best BLAST hit taxonomic
assignment}\label{best-blast-hit-taxonomic-assignment}

@@

\subsection{Availability}\label{availability}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Treat data analysis as a software project. This makes for radical
  improvements in \emph{reproducibility}, \emph{reuse},
  \emph{versioning}, \emph{safety}, \emph{automation} and
  \emph{expressiveness}
\item
  input and output data, their locations and type are expressible and
  checked at compile-time using \emph{Datasets} ­3. management of
  dependencies and machine configurations using \emph{Statika}
\item
  automation of AWS cloud resources and processes, including
  distribution and parallelization through the use of \emph{Loquat}
\item
  taxonomic data and related operations are treated natively as what
  they are: graphs, through the use of \emph{Bio4j}
\item
  MG7 provides a sustainable model for taxonomic assignment, appropriate
  to face the challenging amount of data that high throughput sequencing
  technologies generate
\end{enumerate}

We will expand on each item in the following sections.

@@

@@

We use EC2, S3 and SQS through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK},
\href{https://github.com/ohnosequences/aws-scala-tools/???}{ohnosequences/aws-scala-tools
0.13.2}. This uses version \texttt{1.10.9} of the AWS Java SDK.

MG7 uses
\href{https://github.com/statika/statika/releases/tag/v0.2.0}{ohnosequences/statika
2.0.0} for specifying the configuration and behavior of EC2 instances.

MG7 uses
\href{https://github.com/ohnosequences/datasets/releases/tag/v0.2.0}{ohnosequences/datasets
0.2.0} for specifying input and output data, their type and their
location.

MG7 uses
\href{https://github.com/ohnosequences/loquat/releases/tag/v2.0.0}{ohnosequences/loquat
2.0.0} for the specification of data processing tasks and their
execution using AWS resources.

\subsection{BLAST eDSL}\label{blast-edsl}

MG7 uses
\href{https://github.com/ohnosequences/blast/releases/tag/v0.2.0}{ohnosequences/blast
0.2.0}. The BLAST version used is \texttt{2.2.31+}

\subsection{FLASH eDSL}\label{flash-edsl}

MG7 uses
\href{https://github.com/ohnosequences/flash/releases/tag/v0.1.0}{ohnosequences/flash
0.1.0}. The FLASH version used is \texttt{?.?.?}

MG7 uses
\href{https://github.com/bio4j/bio4j/releases/tag/v0.12.0-RC3}{bio4j/bio4j
0.12.0-RC3} and
\href{https://github.com/bio4j/bio4j-titan/releases/tag/v0.4.0-RC2}{bio4j/bio4j-titan
0.4.0-RC2} as an API for the NCBI taxonomy.

@@

The two first authors are funded by INTERCROSSING (Grant 289974).
